---
title: "Chapter 2"
output: html_document
---

# Data analysis exercise
## Data description

Data aims to model the relationship between learning approaches and students achievements in an introductory statistics course in Finland. There are total of 7 variables which explain the students achievements (points). Variables are "gender", "Age", "Attitude","Deep","Stra" and "Surf".

### Data dimensions

Data has 166 observations (students) and 7 variables. Hence, data is a table with 166 rows and 7 columns.

>Variables:

- Age;      Age (in years) derived from the date of birth

- Attitude; Global attitude toward statistics

- Points;   Exam points

- gender;   Gender: M (Male), F (Female)

- Deep;     Deep approach             ~d_sm+d_ri+d_ue where, 
                                      d_sm = Seeking Meaning
                                      d_ri = Relating Ideas
                                      d_ue = Use of Evidence
                                     
- Surf;     Surface approach          ~su_lp+su_um+su_sb where,
                                      su_lp = Lack of Purpose
                                      su_um = Unrelated Memorising
                                      su_sb = Syllabus-boundness

- Stra;     Strategic approach        ~st_os+st_tm where,
                                      st_os = Organized Studying
                                      st_tm = Time Management



### Variable statistics
<br>

```{r students}
url = "http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt"
data_stud <- read.delim(url, sep=",")
summary(data_stud)
```

## Data Visualization
### Plots
```{r students2}
url = "http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt"
data_stud <- read.delim(url, sep=",")
boxplot(split(data_stud$points,data_stud$gender),main='Points by Gender')
boxplot(split(data_stud$attitude,data_stud$gender),main='Attitude by Gender')
boxplot(split(data_stud$deep,data_stud$gender),main='Deep by Gender')
boxplot(split(data_stud$stra,data_stud$gender),main='Stra by Gender')
boxplot(split(data_stud$surf,data_stud$gender),main='Surf by Gender')
pairs(~points+age+attitude+deep+stra+surf,data=data_stud, 
   main="Scatterplot Matrix")
library(corrplot)
data_stud$gender <- factor(data_stud$gender)
num_data <- data_stud[,c("points","age","attitude","deep","stra","surf")]
res <- cor(num_data)
corrplot(as.matrix(res), method="circle")
```

<br>
<br>

# Comments on data

### Gender
There seems to be slight variation in points between the genders:
```{r students3}
url = "http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt"
data_stud <- read.delim(url, sep=",")
tapply(data_stud$points, data_stud$gender, mean)
```
Difference in mean points is (23.48214-22.32727) = 1.16

**Conclusion:** males tend to get slightly higher points than women.

Interesting feature between genders is that especially in attitude-variable (which describes general attitude towards statistics) is higher in males than females. Females on the other hand, are higher at stra (strategic approach: organizing studies and time management). Females might find lack of purpose in statistic related studies and studying it might include unrealted memorising because of it (Surf). 


### Age

When looking at age-variable, one should note that there is extremely high number of occurences 20-30-years category: it might give us bad image from how age truly affects the total points of student.

Correlation between points and age:

```{r students4}
url = "http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt"
data_stud <- read.delim(url, sep=",")
with(data_stud, cor(age, points))
```

Correlation is negative.

**Conclusion:** older the student is - lower the points are. Although, correlation is not particularly strong.

### Attitude, Stra, Deep and Surf

Correlations between students attitude and study habits, and exam points.

```{r students5}
url = "http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt"
data_stud <- read.delim(url, sep=",")
with(data_stud, cor(attitude, points))
with(data_stud, cor(stra, points))
with(data_stud, cor(deep, points))
with(data_stud, cor(surf, points))
```

The best variable which indicates students points is, by far, **attitude**. Correlation is 0.44.

Stra and Surf are give second best interpretiations: Students with strategic approach tend to get better points, on the other hand Surf decreases students points (Lack of Purpose, Unrelated Memorising, Syllabus-boundness).


## Correlations between the descriptive variables

From correlation plot (with circles from red - blue), we are able to notice that Surf and Deep has strong negative correlation. Students that lack motivation, do not seek new ideas or meaning from statistics.

Strategic approach (Stra) correlates positively between all the variables, except Surf. 


*Describe the work you have done this week and summarize your learning.*



- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using 

<br>
<br>

# Fitting regression model

## Model parameters

I chose three explanatory variables which are:

- Attitude
- Stra
- Age

```{r students6}
url = "http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt"
l_model <- lm(points ~ attitude + stra + age, data=data_stud)
summary(l_model)

```

### What it does?

Multiple linear regression explains the relationship between descriptive variables (attitude, stra, age) and response variables (points). Model fits linear (linear line, bluntly) equation to observed data.

### Results

As expected, attitude shows strong statistical significance in the model output. Attitudes p-value is clearly under 0.001 (4.7*10^-9, to be exact). Strategic approach (Stra) shows clear statistical significance: altough p-value is just a bit over 0.05, it is still rather excellent explanator for students points. Age is the worst explanator: p-value of age is high as 0.09.

Although, some of the explanatory variables show statistical significance in linear model, the R-squared value of the whole model stays rather low: 0.20.

Model: 
Points_est = 10.89543 + (3.48077*attitude) +  (1.00371*stra) - (0.08822*age)

<br>
<br>

### Re-model, without age

```{r students7}
url = "http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt"
l_model <- lm(points ~ attitude + stra, data=data_stud)
summary(l_model)

```

### Results without age

Model got worse without age-variable. R-squared in new model 0.19.
Conclusion: continue with attitude, stra and age.

<br>
<br>

## Diagnostic plots

```{r students8}
url = "http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt"
l_model <- lm(points ~ attitude + stra + age, data=data_stud)
plot(l_model)

```

## Diagnostics interpretation

**Aim** of the diagnostics plots is to explain how well our multilinear model is actually fitting the data.

### Residuals vs Fitted

Scatterplot represents the correlation between model residuals and predicted values. If residuals are small, the model predicts well. On the other hand, if the scatterplot seems sparse, it indicates that model is not giving good predictions. Remember, out model R^2 was only 0.2.

### QQ-plot

Normal Q-Q-plot is a normal probability plot: if errors in model follow normal distribution, line is straight. In our case, there seems to be slight divergence from normal distribution in both ends. Extremely low and high points are harder to predict with the model.

### Residuals vs Leverage

Last plot describes the influence of observations in the model. We can interpret from the plot that obs. 56, 4 and 2 have the most influence to our model. We could go back to the original data and check is there something wrong with it, or make conclusion why these particular students have such a different point - attitude - stra - and age combinations. They are not clearly following the trend.

